---
title: "Support Vector Machines"
output:
  html_document: default
  html_notebook: default

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(4321)

packages = c("kernlab", "tm", "e1071", "SnowballC", "caret")

## Now load or install&load all
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)
```

# Introduction to SVM in R
In this practice we are going to apply SVM to synthetic data to understand its operation and the impact of its parameters.
The example has been taken from https://escience.rpi.edu/data/DA/svmbasic_notes.pdf

## Data Generation

We generate some data belonging to two classes linearly separables.
```{r}
n <- 1500          # number of data points
p <- 2             # dimension
sigma <- 1         # variance of the distribution
meanpos <- 0       # centre of the distribution of positive examples
meanneg <- 3       # centre of the distribution of negative examples
npos <- round(n/2) # number of positive examples
nneg <- n-npos     # number of negative examples

# Generate the positive and negative examples
Xpositive <- matrix(rnorm(npos*p,mean=meanpos,sd=sigma), npos, p)
Xnegative <- matrix(rnorm(nneg*p,mean=meanneg,sd=sigma), npos, p)
x <- rbind(Xpositive,Xnegative)

# Generate the labels
y <- matrix(c(rep(1,npos), rep(-1,nneg)))

```

## Data Visualization

Let's visualize the data
```{r}
# Visualize the data
plot(scale(x),col=ifelse(y>0,1,2))
legend("topleft", c('Positive','Negative'), col=seq(2), pch=1, text.col=seq(2), cex=0.8)
```

## Train Test Splitting

For the later evaluation we split the dataset (80% for training and 20% for test)
```{r}
## Prepare a training and a test set ##
ntrain <- round(n*0.8)     # number of training examples
tindex <- sample(n,ntrain) # indices of training samples
xtrain <- x[tindex,]
xtest <- x[-tindex,]
ytrain <- y[tindex]
ytest <- y[-tindex]
istrain = rep(0, n)
istrain[tindex] = 1
# Visualize
plot(scale(x),col=ifelse(y>0,1,2),pch=ifelse(istrain==1,1,2), xlab="X1", ylab="X2")
legend("topleft",c('Positive Train','Positive Test','Negative Train','Negative Test'),
       col=c(1,1,2,2),pch=c(1,2,1,2),text.col=c(1,1,2,2))
```

## SVM
We will make use of the `ksvm` function in the `kernlab` package to train a SVM classifier on the generated data. 

There are two important parameters:  C that controls the penalty of wrongly categorize datapoints depending on their distance from the separating hyperplane.
As we explained in class, selecting a proper C vaule is crucial to the generalizability of the SVM. In this example code, it´s set to 1.

The other parameter is the kernel applied. For this example, we use a linear kernel (called "vanilladot").

```{r message = FALSE, warning=FALSE}
#
# Plot the decision boundary
#
set.seed(101)
Cvalue <- 1         
x <- rbind(matrix(rnorm(120),,2), matrix(rnorm(120,mean=3),,2))
y <- matrix(c(rep(1,60),rep(-1,60)))
svp <- ksvm(x,y,type="C-svc", C=Cvalue, kernel = "vanilladot")
plot(svp,data=x)
#
## Explanation for plotting decision boundary can be found at:
##   http://www.di.fc.ul.pt/~jpn/r/svm/svm.html
#
```

```{r}
# Predict labels on test
ypred = predict(svp,xtest)

print("Confusion Matrix:")
table(ytest,ypred)

# Compute accuracy
print("Accuracy:")
sum(ypred==ytest)/length(ytest)

```

# Exercise 1: RBF Kernel

Based on the previous code, test a non-linear kernel and plot the decision boundary.
¿How does it affect to the classification boundary?
```{r}
set.seed(101)
Cvalue <- 100   
x <- rbind(matrix(rnorm(120),,2), matrix(rnorm(120,mean=3),,2))
y <- matrix(c(rep(1,60),rep(-1,60)))
# Your Code here
```


```{r}
# Predict labels on test

# Compute accuracy

```

# Exercise #2: Regularization

Play with different values of `C` and explain how they affect to the decision boundary.

```{r}

set.seed(101)
Cvalue <- 0.01  
x <- rbind(matrix(rnorm(120),,2), matrix(rnorm(120,mean=3),,2))
y <- matrix(c(rep(1,60),rep(-1,60)))

# Your Code here
```


```{r}

set.seed(101)
Cvalue <- 1000       
x <- rbind(matrix(rnorm(120),,2), matrix(rnorm(120,mean=3),,2))
y <- matrix(c(rep(1,60),rep(-1,60)))

# Your Code here
```



# Exercise #3: Polynomial Kernel

We have played with linearly separable data. Let's see now how SVM behaves for non-linearly separable data. To that end we load the "spiral data", where each color belongs to one class.
```{r}
train <- read.csv("datasets/train.csv")
test <- read.csv("datasets/test.csv")

# color list for each of the classes. The example has three classes but one extra is needed as required by filled.contour
clist      <- rainbow(4, s = 1,   v = 1, start = 0.2, 1, alpha = 1)
clistdesat <- rainbow(4, s = 0.5, v = 1, start = 0.2, 1, alpha = 1)

for (c in 1:3) {
	train$color[train$Label==c-1] <- clist[c];
	train$colordesat[train$Label==c-1] <- clistdesat[c];
}

# plot train data
plot(train$X, train$Y, bg=train$color , pch=21,  main="Spiral Train", xlab="X", ylab="Y");

#  transform Label column to factor so svm treats it as a categorical classification
train$Label <- factor(train$Label, levels=c("0", "1", "2"), ordered=T)
test.label <- factor(train$Label, levels=c("0", "1", "2"), ordered=T)

```

For this Exercise we will use the `e1071` library.
The following function train a SVM model with a polynomial kernel and plot the classification boundary

```{r}

# load svm library 
library(e1071);
svm_model <- function(title="SVM", kernel='polynomial', degree=3) {
	model <- svm(Label ~ X + Y, data = train, kernel=kernel, degree=degree);
	print(model);
	summary(model);
	plot_title = paste(title,'kernel',kernel,'degree', toString(degree));

# predict and adjust offset so Labels return to the original set of {0, 1, 2}
	pred <- as.numeric(predict(model, test)) - 1;

# reshape prediction vector to a matrix to produce decision boundaries and areas
	z <- matrix(pred, nrow = sqrt(nrow(test)), byrow=F)
# Make a contour plot for the decision boundaries. Also called "level plot"
		filled.contour(x = seq(-1, 1, length.out = nrow(z)),
				y = seq(-1, 1, length.out = ncol(z)),
				z, levels = seq(-0.5, 3 , 1), col = clistdesat,
				xlab="X", ylab="Y", main=plot_title, plot.axes = {points(train$X, train$Y, bg=train$color, pch=21); axis(1); axis(2) });
}

```


If we set the degree of the polynomial kernel to 1, it behaves like a linear SVM (without any kernel).
```{r}
svm_model(degree=1)
```

As you can see, the Linear Kernel is not able to classify the non-linearities in the data.



Try different degrees for the polynomial kernel using the `svm_model` function.
How does it affect the decission boundary? 
Which one would you select?
```{r}
# Your Code here
```

# Exercise #4: Working Example - Text Classification

We are going to apply SVM to a real scenario.
Do you remember the polarity classification task that we addresed in the Naïve Bayes Practice? Let's see if the fancier SVM model can outperform our first naïve approach.

To facilitate you the exercise, the following code prepare the the movie-pang dataset (use the same `movie-pang02.csv` provided for the Naïve Bayes practice) for the application of SVM (use the `trainin_data` and `test_data` objects).


```{r, echo=TRUE, message=FALSE, warning=FALSE}

dataset <- read.csv("datasets/movie-pang02.csv", stringsAsFactors = FALSE)

# Randomize the dataset to facilitate the training process
set.seed(123)
dataset <- dataset[sample(nrow(dataset)), ]
dataset <- dataset[sample(nrow(dataset)), ]

# Convert the target variable ('class') from character to factor.
dataset$class <- as.factor(dataset$class)

corpus <- Corpus(VectorSource(dataset$text))


cleanCorpus <- function(corpus) {
  corpus <-tm_map(corpus, stemDocument)
  corpus.tmp <- tm_map(corpus,removePunctuation)
  corpus.tmp <- tm_map(corpus.tmp,stripWhitespace)
  corpus.tmp <- tm_map(corpus.tmp,removeWords,stopwords("en"))
  return(corpus.tmp)
}

corpus.clean <- cleanCorpus(corpus)

dtm <- DocumentTermMatrix(corpus.clean,control = list(weighting= function(x) weightBin(x)))
dtm <- removeSparseTerms(dtm, .99)

dataset.train <- dataset[1:1500,]
dataset.test <- dataset[1501:2000,]

dtm.train <- dtm[1:1500,]
dtm.test <- dtm[1501:2000,]

corpus.clean.train <- corpus.clean[1:1500]
corpus.clean.test <- corpus.clean[1501:2000]

X <- as.matrix(dtm.train)
y <- dataset.train$class

training_data <- as.data.frame(cbind(y,X))
test_data <- as.data.frame(as.matrix(dtm.test))
```


Use the functions in `e1071` package to create an SVM model for the training data
```{r}
# Your code here
```

Evaluate the SVM model in terms of Accuracy
¿Have we improved the Naive Bayes model?

Predict and compute the confusion matrix
```{r}
# Your code here
```

Calculate the accuracy from the confusion matrix
```{r}
acc <- function(table){
  TP = table[1,1];  # true positives
  TN = table[2,2];  # true negatives
  FP = table[1,2];  # false positives
  FN = table[2,1];  # false negatives
  acc = (TP + TN)/(TP + TN + FP + FN)
  return(acc)
}

# Your code here

```
As expected, our SVM model is significantly better than the more basic NB model.


Let's try to tune SVM parameters to further improve the model performance by using caret and cross validation
```{r}
# Your code here

```
